Symbolic Weighted Language Models and\\ Quantitative Parsing over Infinite Alphabets

### 2021-07-19 main.tex
Ajout du newpage pour marquer les 12 pages
proof property closure SWA byb SWT

\newcommand{\A}{\mathcal{A}} % automata, VPA
\newcommand{\B}{\mathcal{B}} % automata (in constructions)
\newcommand{\C}{\mathcal{C}} % automata (in constructions)
\newcommand{\D}{\mathcal{D}}  % transducer = distance
\newcommand{\E}{\mathbb{E}}
%\newcommand{\P}{\mathcal{P}}
\newcommand{\Q}{\mathcal{Q}}
\newcommand{\R}{\mathcal{R}}
\newcommand{\T}{\mathcal{T}}
\newcommand{\W}{\mathbb{W}}
\newcommand{\Semiring}{\mathbb{S}}
\newcommand{\zero}{\mathbb{0}}
\newcommand{\one}{\mathbb{1}}
\newcommand{\dom}{\ensuremath{\mathit{dom}}}

\def\SA{\textsf{sA}\xspace}
\def\WA{\textsf{wA}\xspace}
\def\SW{\textsf{sw}\xspace}
\def\swM{\textsf{swM}\xspace}
\def\SWT{\textsf{swT}\xspace}
\def\SWA{\textsf{swA}\xspace}
\def\SWTA{\textsf{swTA}\xspace}
\def\SWVPA{\textsf{sw-VPA}\xspace}
\def\VPA{\textsf{VPA}\xspace}
\def\SVPA{\textsf{sVPA}\xspace}
\def\weight{\mathsf{weight}}
\def\wei{\mathsf{w}}
\def\mei{\mathsf{m}}
\def\init{\mathsf{in}}
\def\final{\mathsf{out}}
\def\Omegai{{\Omega_\mathsf{i}}}
\def\Omegac{{\Omega_\mathsf{c}}}
\def\Omegar{{\Omega_\mathsf{r}}}
\def\Sigmai{{\Sigma_\mathsf{i}}}
\def\Sigmac{{\Sigma_\mathsf{c}}}
\def\Sigmar{{\Sigma_\mathsf{r}}}
\def\Deltai{{\Delta_\mathsf{i}}}
\def\Deltac{{\Delta_\mathsf{c}}}
\def\Deltar{{\Delta_\mathsf{r}}}
\def\Phii{{\Phi_\mathsf{i}}}
\def\Phic{{\Phi_\mathsf{c}}}
\def\Phir{{\Phi_\mathsf{r}}}
\def\Phix{{\Phi_\mathsf{x}}}
\def\Phici{{\Phi_\mathsf{ci}}}
\def\Phicc{{\Phi_\mathsf{cc}}}
\def\Phicr{{\Phi_\mathsf{cr}}}
\def\Phicx{{\Phi_\mathsf{cx}}}
\def\weii{{\wei_\mathsf{i}}}
\def\weic{{\wei_\mathsf{c}}}
\def\weir{{\wei_\mathsf{r}}}
\newcommand{\weie[1]}{\wei_{#1}^\mathsf{e}}
\def\weiei{\weie[\mathsf{i}]}
\def\weiec{\weie[\mathsf{c}]}
\def\weier{\weie[\mathsf{r}]}
\def\weiex{\weie[\mathsf{x}]}
\newcommand{\ioi}[1]{\mathsf{ioi}({#1})}
\newcommand{\rank}{\mathsf{rk}}
\newcommand{\lin}{\mathsf{lin}}
\newcommand{\ts}[1]{\mathop{:}{#1}} % timestamp marker
%\newcommand{\ts}[1]{^{#1}} % timestamp marker

@@

%\paragraph*{Semirings}
%\label{section:semiring}\label{sec:semiring}
\input{semiring}

%\paragraph*{Label Theory}
%\label{section:symbols}
\input{labels}

@@

\lydia{added $u$ and $v$ def}

@@

\begin{example}\label{ex:SWT}
We build a small \SWT over the alphabets $\Sigma$ and $\Deltai$ of
Examples~\ref{ex:running} and \ref{distance-time},
%weighted transducer model
with two states $q_0$ and $q_1$ that
calculates the temporal distance between an input performance in~$\Sigma^*$
and the subsequence
of the events of $\Deltai$ in a score.
%the performance $I$ and
%the score $O$ of Example~\ref{ex:running}.
%
When reading one performed event $\mu$
and the corresponding notated event $\nu$
(for instance MIDI pitch 69 and note name $\mbox{\footnotesize A4}$),
the weight value computed by the \SWT is the time distance between both,
(see Example~\ref{distance-time}), as modeled by transitions $\wei_{11}$ below.
%
If we meet the continuation symbol $'-'$
(this occurs for instance in \emph{ties}
$\musQuarter\!\!\!\mathrel{\raisebox{-1.5mm}{$\smile$}}\!\!\!\musEighth$,
or \emph{dots} \musQuarterDotted{}),
%that represents continuation
it is  skipped with no cost (transitions $\wei_{01}$ or weight $\one$).
%Likewise, the markups are ignored.
\[
\begin{array}{lclclcll}
\wei_{11}(q_0, \mu\ts{\tau}, \nu\ts{\tau'}, q_0) & = &
     \delta(\mu\ts{\tau}, \nu\ts{\tau'}) & \quad & %|\tau' - \tau| & \quad &
\wei_{11}(q_1, \mu\ts{\tau}, \nu\ts{\tau'}, q_0) & = &
     \delta(\mu\ts{\tau}, \nu\ts{\tau'}) &
     \mbox{if~} \nu \neq - \\ %|\tau' - \tau| & \quad &
\wei_{01}(q_0, \varepsilon, \mathsf{-}\ts{\tau'}, q_0) & = & \one & &
\wei_{01}(q_1, \varepsilon, \mathsf{-}\ts{\tau'}, q_0) & = & \one\\
\wei_{10}(q_0, \mu\ts{\tau}, \varepsilon, q_1) & = & \alpha & & %\multicolumn{3}{l}{\mathrm{for~all~} b \in \Delta}
\end{array}
\]
%
We also want to take performing errors into account,
since a performer could, for example, play an unwritten extra note.
%\lydia{reformulated this sentence}
%
%Reading such a note is modelled by
The transition $\wei_{10}$,
with an fixed weight value $\alpha \in \Semiring$,
switches from state $q_0$ (normal) to $q_1$ (error)
when reading an extra note $\mu$.
\philippe{Comprends pas cette phrase}
\florent{$\to$ reformulation}
The transitions in the second column below switch back to the normal state $q_0$.
% the metric computed by the \SWT is the smallest sum of point wise distances
% between dates of input and output events.
At last, we let $q_0$ be the only initial and final state, with
$\init(q_0) = \final(q_0) = \one$, and
$\init(q_1) = \final(q_1) = \zero$.
%\florent{dernière phrase utile?}
%This \SWT is capable of evaluating the differences between a score and a performance,
%all the while ensuring that performance errors are plausible.
%$\init(q_0, d, b) = \final(q_0, d, b) = \one$, and
%$\init(q_1, d, b) = \final(q_1, d, b) = \zero$,
%for all $d \in \Sigma$ and $b \in \Delta$.
\endex
\end{example}

@@

\begin{proposition} \label{prop:epsilon}
Given a \SWT $T$ over $\Sigma$, $\Delta$,
$\Semiring$ commutative, bounded and complete,
and $\bar\Phi$ effective,
and a $\SWA$ $A$ over $\Sigma$, $\Semiring$ and $\bar\Phi$,
there exists an effectively constructible \SWA
$B_{A, T}$ over $\Delta$, $\Semiring$ and $\bar\Phi$,
such that for all $t \in \Delta^*$,
$B_{A, T}(t) = \displaystyle\bigoplus_{s\in \Sigma^*} A(s) \otimes T(s, t)$.
\end{proposition}
%
\begin{proof}
(sketch, see Appendix~\ref{sec:closure} for details).
The state set of $B_{A, T}$ is the Cartesian product of
the state sets of $A$ and $T$ and its transitions simulate,
while reading an output word $t \in \Delta^*$, the
synchronized behaviour of $A$ and $T$ on $t$
and some input word $s \in \Sigma^*$.
The weight for reading of the input is obtained with
$\bigoplus^1_\Sigma$.
%
The main difficulty comes from the transitions
of the form $\wei_{10}$, which read in input and ignore the output.
Such transition shall be simulated by $\varepsilon$-transitions in $B_{A, T}$,
but $\varepsilon$-transitions are not defined for $\SWA$.
Therefore, the $\varepsilon$-transitions
are eliminated on-the-fly during the construction of~$B_{A, T}$,
following a procedure of~\cite{LombardySakarovitch12ciaa}.
%\qed
\end{proof}

%The complexity of construction
\noindent
The construction time and size for $B_{A, T}$ are $O(\| T \|^3 . \| A \|^2)$,
where the sizes $\| T \|$ and $\| A \|$ are their number of states.
%of~$T$ is its number of states $|Q|$.
\florent{revise complexity with nb. of tr. and states}
%
The particular case of Proposition~\ref{prop:epsilon} with a singleton $A$,
\ie such that $A(s) = \one$ for a given $s \in \Sigma^*$
and $A(s') = \zero$ for all $s' \neq s$,
corresponds to a construction of a \SWA for the partial application of the \SWT $T$,
fixing the first argument $s$.
\begin{corollary} \label{cor:epsilon}
Given a \SWT $T$ over $\Sigma$, $\Delta$,
$\Semiring$ commutative, bounded and complete,
and $\bar\Phi$ effective,
and $s \in \Sigma^+$,
there exists an effectively constructible \SWA
$B_{s, T}$ over $\Delta$, $\Semiring$ and $\bar\Phi$,
such that for all $t \in \Delta^*$, $B_{s, T}(t) = T(s, t)$.
\end{corollary}

@@

\section{SW Visibly Pushdown Automata}
\label{section:SWVPA}\label{sec:SWVPA}
The model presented in this section generalizes symbolic \VPA
(\SVPA~\cite{dAntonyAlur14SVPDA},
 generalizing themselves \VPA~\cite{AlurMadhusudan09nested} to infinite alphabets)
from Boolean semirings to arbitrary semiring weight domains.
It will compute on nested words over infinite alphabets,
associating to every such word a weight value.
Nested words are able to describe structures of labeled trees,
and in the context of parsing, they will be useful to
represent AST. % parse trees.
\florent{see \S 5 and App.A}

@@

\section{Nested-Words and Parse-Trees} \label{sec:trees}
\input{app-trees}

>>>>

\newcommand{\A}{\mathcal{A}} % automata, VPA
\newcommand{\B}{\mathcal{B}} % automata (in constructions)
\newcommand{\C}{\mathcal{C}} % automata (in constructions)
\newcommand{\D}{\mathcal{D}}  % transducer = distance
\newcommand{\E}{\mathbb{E}}
%\newcommand{\P}{\mathcal{P}}
\newcommand{\Q}{\mathcal{Q}}
\newcommand{\R}{\mathcal{R}}
\newcommand{\T}{\mathcal{T}}
\newcommand{\W}{\mathbb{W}}
\newcommand{\Semiring}{\mathbb{S}}
\newcommand{\zero}{\mathbb{0}}
\newcommand{\one}{\mathbb{1}}
\newcommand{\dom}{\ensuremath{\mathit{dom}}}
\newcommand{\seq}{\ensuremath{\mathit{seq}}}
\newcommand{\word}{\mathit{word}}

\def\SA{\textsf{sA}\xspace}
\def\WA{\textsf{wA}\xspace}
\def\SW{\textsf{sw}\xspace}
\def\swM{\textsf{swM}\xspace}
\def\SWT{\textsf{swT}\xspace}
\def\SWA{\textsf{swA}\xspace}
\def\SWTA{\textsf{swTA}\xspace}
\def\SWVPA{\textsf{sw-VPA}\xspace}
\def\VPA{\textsf{VPA}\xspace}
\def\SVPA{\textsf{sVPA}\xspace}
\def\weight{\mathsf{weight}}
\def\mei{\mathsf{m}}
\def\init{\mathsf{in}}
\def\final{\mathsf{out}}
\def\Omegai{{\Omega_\mathsf{i}}}
\def\Omegac{{\Omega_\mathsf{c}}}
\def\Omegar{{\Omega_\mathsf{r}}}
\def\Sigmai{{\Sigma_\mathsf{i}}}
\def\Sigmac{{\Sigma_\mathsf{c}}}
\def\Sigmar{{\Sigma_\mathsf{r}}}
\def\Deltai{{\Delta_\mathsf{i}}}
\def\Deltac{{\Delta_\mathsf{c}}}
\def\Deltar{{\Delta_\mathsf{r}}}
\def\Phii{{\Phi_\mathsf{i}}}
\def\Phic{{\Phi_\mathsf{c}}}
\def\Phir{{\Phi_\mathsf{r}}}
\def\Phix{{\Phi_\mathsf{x}}}
\def\Phici{{\Phi_\mathsf{ci}}}
\def\Phicc{{\Phi_\mathsf{cc}}}
\def\Phicr{{\Phi_\mathsf{cr}}}
\def\Phicx{{\Phi_\mathsf{cx}}}
\def\wei{\mathsf{w}}
\def\weii{{\wei_\mathsf{i}}}
\def\weic{{\wei_\mathsf{c}}}
\def\weir{{\wei_\mathsf{r}}}
\newcommand{\weie[1]}{\wei_{#1}^\mathsf{e}}
\def\weiei{\weie[\mathsf{i}]}
\def\weiec{\weie[\mathsf{c}]}
\def\weier{\weie[\mathsf{r}]}
\def\weiex{\weie[\mathsf{x}]}
\newcommand{\ioi}[1]{\mathsf{ioi}({#1})}
\newcommand{\rank}{\mathsf{rk}}
\newcommand{\lin}{\mathsf{lin}}
\newcommand{\ts}[1]{\mathop{:}{#1}} % timestamp marker
%\newcommand{\ts}[1]{^{#1}} % timestamp marker

@@

%\paragraph*{Semirings}
\label{section:semiring}\label{sec:semiring}
\input{semiring}

%\paragraph*{Label Theory}
\label{section:symbols}
\input{labels}

@@

%\lydia{added $u$ and $v$ def}

@@

\begin{example}\label{ex:SWT}
We build a small \SWT over the alphabets $\Sigma$ and $\Deltai$ of
Examples~\ref{ex:running} and \ref{distance-time},
%weighted transducer model
with two states $q_0$ and $q_1$ that
calculates the temporal distance between an input performance in~$\Sigma^*$
and the subsequence
of the events of $\Deltai$ in a score.
%the performance $I$ and
%the score $O$ of Example~\ref{ex:running}.
%
When reading one performed event $\mu$
and the corresponding notated event $\nu$
(for instance MIDI pitch 69 and note name $\mbox{\footnotesize A4}$),
the weight value computed by the \SWT is the time distance between both,
(see Example~\ref{distance-time}), as modeled by transitions $\wei_{11}$ below.
%
If we meet the continuation symbol $'-'$
(this occurs for instance in \emph{ties}
$\musQuarter\!\!\!\mathrel{\raisebox{-1.5mm}{$\smile$}}\!\!\!\musEighth$,
or \emph{dots} \musQuarterDotted{}),
%that represents continuation
it is  skipped with no cost (transitions $\wei_{01}$ or weight $\one$).
%Likewise, the markups are ignored.
\[
\begin{array}{lclclcll}
\wei_{11}(q_0, \mu\ts{\tau}, \nu\ts{\tau'}, q_0) & = &
     \delta(\mu\ts{\tau}, \nu\ts{\tau'}) & \quad & %|\tau' - \tau| & \quad &
\wei_{11}(q_1, \mu\ts{\tau}, \nu\ts{\tau'}, q_0) & = &
     \delta(\mu\ts{\tau}, \nu\ts{\tau'}) &
     \mbox{if~} \nu \neq - \\ %|\tau' - \tau| & \quad &
\wei_{01}(q_0, \varepsilon, \mathsf{-}\ts{\tau'}, q_0) & = & \one & &
\wei_{01}(q_1, \varepsilon, \mathsf{-}\ts{\tau'}, q_0) & = & \one\\
\wei_{10}(q_0, \mu\ts{\tau}, \varepsilon, q_1) & = & \alpha & & %\multicolumn{3}{l}{\mathrm{for~all~} b \in \Delta}
\end{array}
\]
%
We also want to take performing errors into account,
since a performer could, for example, play an unwritten extra note.
%\lydia{reformulated this sentence}
%
%Reading such a note is modelled by
The transition $\wei_{10}$,
with an fixed weight value $\alpha \in \Semiring$,
switches from state $q_0$ (normal) to $q_1$ (error)
when reading an extra note $\mu$.
%\philippe{Comprends pas cette phrase}
%\florent{$\to$ reformulation}
The transitions in the second column below switch back to the normal state $q_0$.
% the metric computed by the \SWT is the smallest sum of point wise distances
% between dates of input and output events.
At last, we let $q_0$ be the only initial and final state, with
$\init(q_0) = \final(q_0) = \one$, and
$\init(q_1) = \final(q_1) = \zero$.
%\florent{dernière phrase utile?}
%This \SWT is capable of evaluating the differences between a score and a performance,
%all the while ensuring that performance errors are plausible.
%$\init(q_0, d, b) = \final(q_0, d, b) = \one$, and
%$\init(q_1, d, b) = \final(q_1, d, b) = \zero$,
%for all $d \in \Sigma$ and $b \in \Delta$.
\endex
\end{example}

@@

\begin{proposition} \label{prop:epsilon}
Given a \SWT $T$ over $\Sigma$, $\Delta$,
$\Semiring$ commutative, bounded and complete,
and $\bar\Phi$ effective,
and a $\SWA$ $A$ over $\Sigma$, $\Semiring$ and $\bar\Phi$,
there exists an effectively constructible \SWA
$B_{A, T}$ over $\Delta$, $\Semiring$ and $\bar\Phi$,
such that for all $t \in \Delta^+$,
$B_{A, T}(t) = \displaystyle\bigoplus_{s\in \Sigma^*} A(s) \otimes T(s, t)$.
\end{proposition}
%
\begin{proof}
(sketch, see Appendix~\ref{sec:closure} for details).
The state set of $B_{A, T}$ is the Cartesian product of
the state sets of $A$ and $T$ and its transitions simulate,
while reading an output word $t \in \Delta^*$, the
synchronized behaviour of $A$ and $T$ on $t$
and some input word $s \in \Sigma^*$.
The weight for reading the input $s$ is obtained with
$\bigoplus^1_\Sigma$.
%
The main difficulty comes from the transitions
of the form $\wei_{10}$, which read in input and ignore the output.
Such transition shall be simulated by $\varepsilon$-transitions in $B_{A, T}$,
but $\varepsilon$-transitions are not defined for $\SWA$.
Therefore, the $\varepsilon$-transitions
are eliminated on-the-fly during the construction of~$B_{A, T}$,
following a procedure of~\cite{LombardySakarovitch12ciaa}.
%\qed
\end{proof}

%The complexity of construction
\noindent
The construction time and size for $B_{A, T}$ are $O(\| T \|^3 . \| A \|^2)$,
where the sizes $\| T \|$ and $\| A \|$ are their number of states.
%of~$T$ is its number of states $|Q|$.
\florent{revise complexity with nb. of tr. and states? 
         or skip details and just say polynomial?}
%
The particular case of Proposition~\ref{prop:epsilon} with a singleton $A$,
\ie such that $A(s) = \one$ for a given $s \in \Sigma^*$
and $A(s') = \zero$ for all $s' \neq s$,
corresponds to a construction of a \SWA for the partial application of the \SWT $T$,
fixing the first argument $s$.
\begin{corollary} \label{cor:epsilon}
Given a \SWT $T$ over $\Sigma$, $\Delta$,
$\Semiring$ commutative, bounded and complete,
and $\bar\Phi$ effective,
and $s \in \Sigma^+$,
there exists an effectively constructible \SWA
$B_{s, T}$ over $\Delta$, $\Semiring$ and $\bar\Phi$,
such that for all $t \in \Delta^+$, $B_{s, T}(t) = T(s, t)$.
\end{corollary}

@@

\section{SW Visibly Pushdown Automata}
\label{section:SWVPA}\label{sec:SWVPA}
The model presented in this section generalizes symbolic \VPA
(\SVPA~\cite{dAntonyAlur14SVPDA},
 generalizing themselves \VPA~\cite{AlurMadhusudan09nested} to infinite alphabets)
from Boolean semirings to arbitrary semiring weight domains.
It will compute on nested words over infinite alphabets,
associating to every such word a weight value.
Nested words are able to describe structures of labeled trees,
and in the context of parsing, they will be useful to
represent AST 
(see Section~\ref{sec:parsing} and Appendix~\ref{app:trees}). % parse trees. 
%\florent{see \S 5 and App.A}

@@

\section{Nested-Words and Parse-Trees} \label{sec:trees}\label{app:trees}
\input{app-trees}
